\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-20 paper.qmd
Update with mse

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. Analysts often turn to NLDR techniques to verify and identify these clusters and explore developmental trajectories. To illustrate the importance of NLDR techniques and parameter selection in identifying clusters, Human Peripheral Blood Mononclear Cells (PBMC3k) dataset [@chen2023] is used. In a study by @chen2023, this dataset was used to demonstrate how UMAP represents clusters (see @fig-umap-author). As shown in @fig-umap-author, there are three distinct and well-separated clusters.

@@

<!--bin1 = 30, bin2 = 29, b = 870, non_empty = 132-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with UMAP, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distinct clusters. The $p\\text{-}D$ model errors are distributed along clusters."
#| label: fig-model-pbmc-author
#| fig-pos: H

@@

<!--bin1 = 15, bin2 = 20, b = 300, non_empty = 137-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with tSNE, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distinct clusters. The $p\\text{-}D$ model errors are distributed along clusters, but most low $p\\text{-}D$ model errors present in the large cluster."
#| label: fig-model-pbmc
#| fig-pos: H
#| out-height: 30%

>>>>

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. Analysts often turn to NLDR techniques to verify and identify these clusters and explore developmental trajectories. To illustrate the importance of NLDR techniques and parameter selection in identifying clusters, Human Peripheral Blood Mononclear Cells (PBMC3k) dataset [@chen2023] is used. In a study by @chen2023, this dataset was used to demonstrate how UMAP represents clusters (see @fig-umap-author). As shown in @fig-umap-author, there are three distant and well-separated clusters.

@@

<!--bin1 = 30, bin2 = 29, b = 870, non_empty = 132-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with UMAP, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distant clusters. The $p\\text{-}D$ model errors are distributed along clusters."
#| label: fig-model-pbmc-author
#| fig-pos: H

@@

<!--bin1 = 15, bin2 = 20, b = 300, non_empty = 137-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with tSNE, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distant clusters. The $p\\text{-}D$ model errors are distributed along clusters, but most low $p\\text{-}D$ model errors present in the large cluster."
#| label: fig-model-pbmc
#| fig-pos: H
#| out-height: 30%

### 2024-05-20 paper.tex
Added removing long edge section
Update with mse

Creating a smooth 2D representation (see \textbf{?@fig-modelScurve} (c))
requires removing edges that connect distant bin centroids within the
triangular mesh (see \textbf{?@fig-modelScurve} (b)). These edges only
exist in the 2D model and do not extend into higher dimensions, ensuring
that their removal does not impact the model in higher dimensions. While
specific criteria for determining the benchmark value to remove long
edges do not exist, the following steps provide an approach to
identifying a suitable threshold:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the 2D Euclidean distances (see
  \textbf{?@fig-distScurve}).
\item
  Identify the largest difference between consecutive distance values.
\item
  Take the largest distance value corresponding to this difference as
  the benchmark value.
\end{enumerate}

@@

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
Analysts often turn to NLDR techniques to verify and identify these
clusters and explore developmental trajectories. To illustrate the
importance of NLDR techniques and parameter selection in identifying
clusters, Human Peripheral Blood Mononclear Cells (PBMC3k) dataset
\citep{chen2023} is used. In a study by \citet{chen2023}, this dataset
was used to demonstrate how UMAP represents clusters (see
Figure~\ref{fig-umap-author}). As shown in Figure~\ref{fig-umap-author},
there are three distinct and well-separated clusters.

@@

\caption{\label{fig-model-pbmc-author}(a) Model generated in
\(2\text{-}D\) with UMAP, and (b) \(p\text{-}D\) model error in
\(2\text{-}D\). The \(2\text{-}D\) model shows three well-separated
distinct clusters. The \(p\text{-}D\) model errors are distributed along
clusters.}

@@

\caption{\label{fig-model-pbmc}(a) Model generated in \(2\text{-}D\)
with tSNE, and (b) \(p\text{-}D\) model error in \(2\text{-}D\). The
\(2\text{-}D\) model shows three well-separated distinct clusters. The
\(p\text{-}D\) model errors are distributed along clusters, but most low
\(p\text{-}D\) model errors present in the large cluster.}

>>>>

To create a smooth \gD{} representation (see
Figure~\ref{fig-NLDR-scurve} (d)), it is necessary to remove edges that
connect distant bin centroids in the triangular mesh. These edges only
exist in the \gD{} model and do not extend into \pD{}, so their removal
does not impact the model in \pD{}. Although there are no specific
criteria for determining the benchmark value to remove long edges, the
following steps provide an approach to identifying a suitable threshold:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the 2D Euclidean distances (see
  Figure~\ref{fig-dist-scurve}).
\item
  Identify the first largest difference between consecutive distance
  values.
\item
  Take the distance value corresponding to this difference as the
  benchmark value.
\end{enumerate}

@@

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
Analysts often turn to NLDR techniques to verify and identify these
clusters and explore developmental trajectories. To illustrate the
importance of NLDR techniques and parameter selection in identifying
clusters, Human Peripheral Blood Mononclear Cells (PBMC3k) dataset
\citep{chen2023} is used. In a study by \citet{chen2023}, this dataset
was used to demonstrate how UMAP represents clusters (see
Figure~\ref{fig-umap-author}). As shown in Figure~\ref{fig-umap-author},
there are three distant and well-separated clusters.

@@

\caption{\label{fig-model-pbmc-author}(a) Model generated in
\(2\text{-}D\) with UMAP, and (b) \(p\text{-}D\) model error in
\(2\text{-}D\). The \(2\text{-}D\) model shows three well-separated
distant clusters. The \(p\text{-}D\) model errors are distributed along
clusters.}

@@

\caption{\label{fig-model-pbmc}(a) Model generated in \(2\text{-}D\)
with tSNE, and (b) \(p\text{-}D\) model error in \(2\text{-}D\). The
\(2\text{-}D\) model shows three well-separated distant clusters. The
\(p\text{-}D\) model errors are distributed along clusters, but most low
\(p\text{-}D\) model errors present in the large cluster.}
